Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2010-11-07 rjournal/RJtemplate.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@21 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{
   A general theory for extending nonparametric goodness-of-fit tests to discrete null distributions has existed
   for several decades. Despite this, modern statistical software have generally failed to provide this methodology
   to users. The packages ks.test and cvm.test serve to fill this need in the R language for
   the two most popular nonparametric tests. The structure of these two packages are explained and examples
   to their specific usage are presented. Particular attention is given to the various numerical issues that arise
   in their implementation.

*** try to clean up, avoid package/function issue, make it a general
contribution if possible.
}

@@

%Why we use non-parametric tests (power to interesting shifts; robustness to underlying assumptions).
While almost any hypothesis test can be viewed as a variant on a goodness-of-fit test, the term is typically
applied to those tests which are nonparametric in nature. That is, they do not conduct the hypothesis testing 
through the explicit calculation of the parameters generating the data. Instead, they generally attempt to 
calculate differences between the empirical distribution of the observed data and the overall distribution 
of the null model. In many cases these tests are preferential since they tend to have an increased power for 
interesting deviations from the null model, in exchange for failing to detect less interesting deviations due to 
factors such as measurement error. By far the most popular of these nonparametric tests are due to Kolmogorov and 
Smirnov, followed closely by several variants on an estimation procedure proposed by Cram\'{e}r-von Mises.

% Discrete versions exist; just no code
While the original aim of nonparametric goodness of fit tests was meant for continuous null distributions, discrete versions
have existed since the early 1970s. Unfortunately, current statistical software has largely failed to incorporate them.
This has left the end-user to either use only parametric tests, such as Pearson's Chi-Squared test, or to incorrectly
apply the functions meant for continuous distributions. As shown in section 5, the latter can be particularly dangerous
in the small sample setting. The packages \pkg{ks.test} and \pkg{cvm.test} serve to fill this void for users
operating in the R environment.

** Ditto on wordsmithing the contribution (package/function, etc...).

@@

Of all the methods for nonparametric goodness-of-fit tests, by far the most popular is the method devised by Kolmogorov and Smirnov.
It is the only such test built into the base of R. The idea behind it is fairly simple. Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution, and the empirical distribution function $F_{data}(x)$ of the observed data, 
one constructs the statistic:
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|
\end{align}
The distribution of $D$ under the null model does not depend on which null distribution is being used, making this a computationally
attractive method. For a standard treatment of the test and its performance relative to other algorithms see \cite{slakter65}. Two common
alternatives of the above test statistic exist. The absolute value is discarded and the rest is either left alone (the `greater' testing 
alternative) or the supremum is replaces with a negative infimum (the `lesser' hypothesis alternative). These can be helpful depending 
on the nature of the alternative hypotheses for which the test is desired to be powerful towards.

@@

Computing the maximum over these $2N$ values (with $f$ equal to $F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic when given a continuous null distribution. When the function $g$ is not
continuous, notice that this formula no longer works since in general we cannot replace the limit of $g$ with its value at $x_i$. If it is known that $g$
is also a step function, we could replace the formula for some small $\epsilon$ by:
\begin{align*}
\sup_x &\left| f(x)- g(x) \right| =   \\
        &                    \max_i \{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \}
\end{align*}
Where the discontinuities in $g$ are at least a distance $\epsilon$ apart. But this requires knowing that $g$ is a step function. In the case of not knowing,
or having a $g$ which is in neither a step function nor a continuous function, the only method for computing the supremum is to take the numerical limit in (1). 
This clearly takes more computational time than simply cycling over $2N$ values.

Therefore, in order to implement the discrete Kolmogorov-Smirnov test statistic, the user must either indicate the points of discontinuity of the null
distribution's cdf or live with a much smaller.

%Once the test statistic is determined, the p-value for this value needs to be computed. For high sample sizes, the test distribution becomes the same as in the
%continuous case. The default options for \pkg{ks.test} already allow the user to specify if they would like approximate asymptotic distributions (since these exist
%in the continuous setting as well). So when this approximation flag is turned on, this distribution is used for all situations. When the user requires exact p-values,
%the methodology in \citet{Conover1972} needs to be used. Code for this procedure in the R language, or in any other open source options, did not previously
%exist and is included in the revised \pkg{ks.test}. The calculations are complex but straightforward; the full details are contained in the package source and in 
%the original Conover paper.

%Numerical issues arise from the summation; need to do something else for high $n$. Simulation? Asymptotic approximation? 
%The interesting part of the implementation of Conover's method, from a computational standpoint, are the difficult numerical issues that arise when calculated the p-values for larger sample sizes.

*** Check Cram\'{e}r-von Mises for hyphen, and let's get the accent done properly.

@@

*** New paragraph, perhaps here, on the differences between these three tests.

@@

%Discuss the discrete version. How to calculate W2, U2, and A2. 
Generalizations of the Cram\'{e}r-von Mises tests were developed in \cite{choulakian1994}. Much like for the Kolmogorov-Smirnov test,
the theoretical form of the test statistics are unchanged; although the discreteness allows for a slightly simpler representation.
The null distribution of the test statistics are again distribution dependent, unlike the continuous version. The methods do not
suggest finite sample results, but rather show that the asymptotic null distribution is equal to a weighted sum of independent
chi-squared variables (the weights depending on the particular distribution). This asymptotic distribution is what we implement here;
the original papers shows that this approximation (*** NOT FINISHED?)

@@

There is no analytic solution to the integral, but integration can be carried out using numerical techniques. This seems
fine in most situations, but numerical issues do become a problem in the regime of large test statistics $x$. The function
$\theta(\cdot, x)$ is linear in $x$, and thus as the the test statistic grows the corresponding period of the integrand 
decreases. As the function acquires too many inflection points, the approximation becomes unstable. This is further
magnified by this occurring when $p$-values should be very small; thus tiny fluctuations which would be undetectable
elsewhere are quite prominent. Figure \ref{cvmissues} shows the non-monotonicity of the function as the test statistic grows.

There is fortunately a simple conservative approximation which can get around this numerical problem. Given the
following inequality:
\begin{align*}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\end{align*}
We see that the values for the weighted sum can be bounded by a simple transformation and a chi-squared distribution
of a higher degree of freedom.

@@

The revised \pkg{ks.test()} function supports one-sample tests for discrete
null distributions by allowing the second argument, \code{y}, to be
an empirical cumulative distribution function (an R function
with class \code{ecdf}) or an object of class \code{stepfun} specifying
a discrete distribution.  As in the original version of \code{ks.test()},
the presence of ties in the data (the first argument, \code{x}) generates a
warning unless \code{y} describes a discrete distribution.  When a discrete
distribution is specified, exact p-values are not available for 
two-sided alternative hypotheses, but the reported p-values will be
conservative.  For one-sided tests,
exact p-values are calculated using Conover's
method (when \code{exact = NULL} or \code{exact = TRUE})
if the sample size is less than or equal to 30; otherwise, asymptotic distributions
are used which are reliable in such cases (CITATION? Is this correct?).
When \code{exact = FALSE} the asymptotic distribution is used which is known
to be imprecise but conservative, even for small sample (CITATION).

@@

Go into more detail about reasons for using these versus chi-squared type tests. (*)

Issues revolving around incorrect usage of default ks.test() in R because of continuity error (i.e. ks.test(1:2, ecdf(1:2)).  Will be earlier, needed here?

Further discuss numerical issues; ks -> n over 30 has problems, CvM -> integrate function can sometimes refuse to run because of failure to calculate integral.
Earlier, needed here?

@@

In the end, for methods such as those presented in this paper where it is possible
there are enough positives to suggest that we attempt to implement exact $p$-value calculations. This does lead
to an interesting direction for future research in non-parametric goodness of fit tests: the methods presented
here were historically chosen because they could be shown to have fixed null distributions which could easily
calculated in an era without fast computers for carrying out permutation tests. It is quite possible that
variants without the property, which could be easily used today, have greater power to certain alternative 
distributions.

Possibly go into a brief discussion of classes and class inheritance in R. The use of `htest' in our functions; benefits and issues with using it. Particularly troublesome in ks.test where we have a range of p-values.  More precisely, explain that we
could provide upper and lower bounds for the p-value but are limited by class
\code{htest}.

Two-sample cases; no theory for discrete distributions. Why? Distribution dependence messes things up. Discuss this for sure. (*)

For further generalizations of tests see \cite{dewev1973}

Future work: continuous CVM.

@@

\address{
  Taylor B. Arnold \\
  24 Hillhouse Ave. \\
  New Haven, CT 06511
  USA\\
}
\email{taylor.arnold@yale.edu}

\address{
  John W. Emerson \\
  24 Hillhouse Ave. \\
  New Haven, CT 06511
  USA\\
}
\email{john.emerson@yale.edu}

>>>>

\abstract{
   A general theory for extending nonparametric goodness-of-fit tests to discrete null distributions has existed
   for several decades. Despite this, modern statistical software have generally failed to provide this methodology
   to users. The packages ks.test and cvm.test serve to fill this need in the R language for
   the two most popular nonparametric tests. The structure of these two packages are explained and examples
   to their specific usage are presented. Particular attention is given to the various numerical issues that arise
   in their implementation.
}

@@

%Why we use non-parametric tests (power to interesting shifts; robustness to underlying assumptions).
While almost any hypothesis test can be viewed as a variant on a goodness-of-fit test, the term is typically
applied to those tests which are nonparametric in nature. That is, they do not conduct the hypothesis testing 
through the explicit calculation of the parameters generating the data. Instead, they generally attempt to 
calculate differences between the empirical distribution of the observed data and the overall distribution 
of the null model. In many cases these tests are preferential since they tend to have an increased power for 
interesting deviations from the null model, in exchange for failing to detect less interesting deviations due to 
factors such as measurement error. By far the most popular of these nonparametric tests are due to Kolmogorov and 
Smirnov, followed closely by several variants on an estimation procedure proposed by Cram\'{e}r and von Mises.

% Discrete versions exist; just no code
While the original aim of nonparametric goodness of fit tests was meant for continuous null distributions, discrete versions
have existed since the early 1970s. Unfortunately, current statistical software has largely failed to incorporate them.
This has left the end-user to either use only parametric tests, such as Pearson's Chi-Squared test, or to incorrectly
apply the functions meant for continuous distributions. As we will see, the latter can be particularly dangerous
in the small sample setting. The functions \pkg{ks.test} and \pkg{cvm.test} presented here serve to fill this void for users
operating in the R environment.

@@

Of all the methods for nonparametric goodness-of-fit tests, the most popular is the method devised by Kolmogorov and Smirnov.
It is the only such test built into the base of R. The idea behind it is fairly simple. Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution, and the empirical distribution function $F_{data}(x)$ of the observed data, 
one constructs the statistic:
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|
\end{align}
The distribution of $D$ under the null model does not depend on which null distribution is being used, making this a computationally
attractive method. For a standard treatment of the test and its performance relative to other algorithms see \cite{slakter65}. Two common
alternatives of the above test statistic exist. The absolute value is discarded and the rest is either left alone (the `greater' testing 
alternative) or the supremum is replaced with a negative infimum (the `lesser' hypothesis alternative). These can be helpful depending 
on the nature of the alternative hypotheses for which the test is desired to be powerful towards.

@@

Computing the maximum over these $2N$ values (with $f$ equal to $F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic when given a continuous null distribution. When the function $g$ is not
continuous, notice that this formula no longer works since in general we cannot replace the limit of $g$ with its value at $x_i$. If it is known that $g$
is also a step function, we could replace the formula for some small $\epsilon$ by:
\begin{align*}
\sup_x &\left| f(x)- g(x) \right| =   \\
        &                    \max_i \{ \left|g(x_i) - f(x_i) \right|, \left| g(x_i - \epsilon) - f(x_{i-1}) \right| \}
\end{align*}
Where the discontinuities in $g$ are at least a distance $\epsilon$ apart. But this requires knowing that $g$ is a step function and knowing something
about its particular break-points. In the case of not knowing,
or having a $g$ which is in neither a step function nor a continuous function, the only method for computing the supremum is to take the numerical limit in (1). 
This clearly takes more computational time than simply cycling over $2N$ values.

Therefore, in order to implement the discrete Kolmogorov-Smirnov test statistic, we have forced the user to indicate the points of discontinuity
of the null distribution's cumulative distribution function. It is often the case that the test is used inside of a long simulation, 
and the added computational time would likely be prohibitive.

Once the test statistic is determined, the p-value for this value needs to be computed. For high sample sizes, the test distribution becomes the same as in the
continuous case. When the user requires exact p-values,
the methodology in \citet{Conover1972} needs to be used. Code for this procedure in the R language, or in any other open source options, did not previously
exist and is included in the new function \pkg{ks.test}. The calculations are complex but straightforward; 
the full details are contained in the package source and in the original Conover paper.

%Numerical issues arise from the summation; need to do something else for high $n$. Simulation? Asymptotic approximation? 
%The interesting part of the implementation of Conover's method, from a computational standpoint, are the difficult numerical issues that 
%arise when calculated the p-values for larger sample sizes.

@@

The relative powers of these tests to different alternatives are studied in depth in \cite{stephens1974}. In general, the $W^2$
statistic is used unless there is a good reason not to. The $A^2$ statistic was developed by Anderson in order to be
generalized to the two-sample case, but was shown to perform very similarly to the original statistic in the one-sample
setting. Watson's $U^2$ statistic was developed for distributions which are cyclically distributed; that is they have some
order to them but no natural starting point. A common example would be the months of the year. His statistic will be same
if the data are cyclically reordered, since giving preference to one ordering would be unnatural.

@@

%Discuss the discrete version. How to calculate W2, U2, and A2. 
Generalizations of the Cram\'{e}r-von Mises tests were developed in \cite{choulakian1994}. Much like for the Kolmogorov-Smirnov test,
the theoretical form of the test statistics are unchanged; although the discreteness allows for a slightly simpler representation.
The null distribution of the test statistics are again distribution dependent, unlike the continuous version. The methods do not
suggest finite sample results, but rather show that the asymptotic null distribution is equal to a weighted sum of independent
chi-squared variables (the weights depending on the particular distribution). This asymptotic distribution is what we implement here;
the original papers shows that this approximation is conservative and asymptotically equivalent to the true null distribution.

@@

There is no analytic solution to the integral, but integration can be carried out using numerical techniques. This seems
fine in most situations, but numerical issues do become a problem in the regime of large test statistics $x$. The function
$\theta(\cdot, x)$ is linear in $x$, and thus as the the test statistic grows the corresponding period of the integrand 
decreases. As the function acquires too many inflection points, the approximation becomes unstable. This is further
magnified by this occurring when $p$-values should be very small; thus tiny fluctuations which would be undetectable
elsewhere are quite prominent.

There is fortunately a simple conservative approximation which can get around this numerical problem. Given the
following inequality:
\begin{align*}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\end{align*}
We see that the values for the weighted sum can be bounded by a simple transformation and a chi-squared distribution
of a higher degree of freedom. While for higher p-values it is better to use the original formulation, for small
p-values (we picked a cut-off of 0.001) this correction is useful and, given the small values, it should not
greatly affect the interpretation of the results.

Figure \ref{cvmissues} shows the non-monotonicity of the function as the test statistic grows, as well as how the
conservative approximation performs. While it is slightly conservative, it has the nice property of being strictly
monotone and not exhibiting other odd noisy behavior.

@@

The revised \pkg{ks.test()} function supports one-sample tests for discrete
null distributions by allowing the second argument, \code{y}, to be
an empirical cumulative distribution function (an R function
with class \code{ecdf}) or an object of class \code{stepfun} specifying
a discrete distribution.  As in the original version of \code{ks.test()},
the presence of ties in the data (the first argument, \code{x}) generates a
warning unless \code{y} describes a discrete distribution.  When a discrete
distribution is specified, exact p-values are not available for 
two-sided alternative hypotheses, but the reported p-values will be
conservative.  For one-sided tests,
exact p-values are calculated using Conover's
method (when \code{exact = NULL} or \code{exact = TRUE})
if the sample size is less than or equal to 30; otherwise, asymptotic distributions
are used which are reliable in such cases (CITATION? Is this correct?).
When \code{exact = FALSE} the asymptotic distribution is used which is known
to be imprecise but conservative, even for small samples (CITATION).

@@

The chi-squared test is a popular alternative to the methods presented here for doing goodness-of-fit tests for discrete data.
It differs by using no information about the geometry of the distribution, instead relying only the difference between the 
number of data points observed at each value and the expected number of data points that should be observed at each point. 
This generally has the effect of having a greater power towards alternatives with sharply different densities on a few points,
at the expense of losing power for alternatives with slight deviations over all of the points. Also, the non-parametric tests
have an advantage of being useful when the sample size is lower compared to the support of the distribution. The chi-squared
test is generally not used unless there are at least five data points in each element of the support. On the other hand,
chi-squared can be used in situations when there is not a numerical interpretation of the observation space 
(e.g. a set of ethnicities).
For a more complete comparison of the two tests' power see \cite{slakter65}.

@@

In the end, for methods such as those presented in this paper where it is possible,
there are enough positives to suggest that we attempt to implement exact $p$-value calculations. This does lead
to an interesting direction for future research in non-parametric goodness of fit tests: the methods presented
here were historically chosen because they could be shown to have fixed null distributions which could easily
calculated in an era without fast computers for carrying out permutation tests. It is quite possible that
variants without the property, which could be easily used today, have greater power to certain alternative 
distributions.

%Possibly go into a brief discussion of classes and class inheritance in R. The use of `htest' in our functions; benefits and issues with using it. Particularly troublesome in ks.test where we have a range of p-values.  More precisely, explain that we
%could provide upper and lower bounds for the p-value but are limited by class
%\code{htest}.

In the continuous setting, both the Kolmogorov-Smirnov and the Cram\'{e}r-von Mises tests have two sample analogues. Here data are
observed from two processes, and the hypothesis tested is whether they came from the same (but unspecified) distribution. There
does not exist an analogous theory for discrete distributions. This comes from the fact that the discrete null distributions
of the test statistics depend on the exact null distribution; therefore the two sample case would surely have to depend
on the exact distributions used as well, which are generally not even stated in the two-sample case.

While we have implemented the two most popular variants of goodness-of-fit tests, there are 
several more exotic varieties to be found. For further generalizations of tests see the extended study done in \cite{dewev1973}.

@@

\address{Taylor B. Arnold \\
Yale University\\
24 Hillhouse Ave. \\
New Haven, CT 06511
USA\\
}
\email{taylor.arnold@yale.edu}

\address{John W. Emerson \\
Yale University\\
24 Hillhouse Ave. \\
New Haven, CT 06511
USA\\
}
\email{john.emerson@yale.edu}
